### 2015-09-15 Manuscript/SW_early_life_telomeres_v9.Rmd
JAE formatting, abstract and intro edited

Lewis G. Spurgin^1,2^,  Kat Bebbington^1^, Eleanor A. Fairfield^1^, Jan Komdeur^3^, Terry Burke^4^, Hannah, L. Dugdale^2^,  and David S. Richardson^1,5^.

1. School of Biological Sciences, University of East Anglia, Norwich Research Park, NR4 7TJ, United Kingdom
2. Department of Zoology, Edward Grey Institute, University of Oxford, Oxford, UK
3. Behavioural Ecology and Self-organization Group, Centre for Ecological and Evolutionary Studies, University of Groningen, Groningen, The Netherlands
4. Department of Animal and Plant Sciences, NERC Biomolecular Analysis Facility, University of Sheffield, Sheffield, UK
5. Nature Seychelles, Roche Caiman, Mahé, Republic of Seychelles

@@

@@

###Abstract
Favourable conditions early in life can confer delayed survival advantages. These 'silver spoon effects' are most likely common in nature and are central to our understanding of life history evolution and population ecology. However, how silver spoon effects vary among individuals, populations and species remains poorly understood. Telomeres can act as biomarkers of cost in wild populations, and may provide insight into how silver spoon effects occur. Using a dataset spanning over a decade, we studied how telomere length links spatiotemporal variation in early-life conditions to survival to in the Seychelles warbler (*Acrocephalus sechellensis*). We show that early-life telomere length varies markedly across birth years, and that this variation is positively related to annual variation in food availability. We then show that having short telomeres as a juvenile is associated with reduced late life survival, but that this relationship was subject to strong cohort effects. Having short telomeres was associated with reduced late-life survival among individuals born in years where food was limited, but not among individuals born in years with high food availability. Our results suggest that telomeres can reveal hidden costs and complex silver spoon effects in wild populations.

###Introduction
Being subject to favourable environmental conditions during development and growth can confer fitness advantages later in life (so called 'silver spoon effects') [@Grafen1988; @Monaghan2008]. There is now evidence that adult fitness can be affected by a range of early-life experiences, including variation in resource availability [@Hayward2013], population density [@Nussey2007] and natural or anthropogenic environmental disturbance [@Reid2003]. Understanding the causes and consequences of these early-life experiences is key to understanding many ecological and evolutionary  processes, including  patterns of natural and sexual selection,  population growth rates and even local extinction [@Coulson2001; @Roach2014]. Understanding silver spoon effects is therefore of central interest to ecologists, evolutionary biologists and conservationists.

@@

The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent opportunity to assess the relationship between early-life environmental variation, telomere length and late-life fitness. Firstly, due to the isolated nature of the study population and intensive field monitoring, we have excellent survival data and sampling material spanning many years (see methods, below). Secondly, telomere length predicts survival independently of age in adult Seychelles warblers, suggesting that telomeres act as a biomarker of cost in this species [@Barrett2013a]. Finally, ecological conditions on Cousin are variable over space and time due to spatiotemporal variation in foliage cover and food availability [@VandeCrommenacker2011a].

@@

###Ethics
All research activities were approved by the ethics committee at the University of East Anglia, the Seychelles Department of Environment and the Seychelles Bureau of Standards (approval reference A0347).

@@

###Competing interests
There are no competing interests

###Authors' contributions
DSR developed the idea and obtained funding for the overall telomere project; DSR, JK, HLD and TB supervised the overall running of the overall Seychelles warbler project and contributed resources; HLD managed the longitudinal fitness database. EAF, KLB and LGS generated and processed the telomere data; LGS performed the analyses and wrote the manuscript, in discussion with DSR. All authors read and contributed to the manuscript.

###Acknowledgements
Nature Seychelles facilitate the long-term Seychelles warbler project, allowing us to stay on Cousin Island, and providing accommodation and facilities during field work. The Seychelles Bureau of Standards and Department of Environment gave permission for sampling and fieldwork. Emma Barrett laid the foundations for this study, generating much of the telomere  data. We thank everyone who has helped in the field, and the current Seychelles warbler research group for useful discussions.

###Funding
This work was funded by two Natural Environment Research Council (NERC) grants to DSR (NE/F02083X/1 and NE/K005502/1) on which JK, TB and HLD were project partners. HLD was funded by a NERC fellowship.

>>>>

Lewis G. Spurgin^a,b^,  Kat Bebbington^a^, Eleanor A. Fairfield^a^, Jan Komdeur^c^, Terry Burke^d^, Hannah, L. Dugdale^c,d^,  and David S. Richardson^a,e,^.

a. School of Biological Sciences, University of East Anglia, Norwich Research Park, NR4 7TJ, United Kingdom
b. Department of Zoology, Edward Grey Institute, University of Oxford, Oxford, UK
c. Behavioural Ecology and Self-organization Group, Centre for Ecological and Evolutionary Studies, University of Groningen, Groningen, The Netherlands
d. Department of Animal and Plant Sciences, NERC Biomolecular Analysis Facility, University of Sheffield, Sheffield, UK
e. Nature Seychelles, Roche Caiman, Mahé, Republic of Seychelles

@@

###Abstract
1. Favourable conditions early in life can confer delayed fitness advantages. However, the mechanistic basis of these 'silver spoon effects' remains poorly understood.  Telomeres - protective caps on the ends of chromosomes - shorten in response to early life stress, and early life telomere shortening can cause reduced later life survival. It is therefore possible that telomeres provide a 'missing link' between early life environmental variation and later life fitness.

2. The prevalence and strength of silver spoon effects has pronounced consequences for individual fitness and population growth rates. Understanding the physiological and ecological links between early life experiences and later life survival is therefore fundamental to our understanding of life-history evolution and population ecology alike.

3. We studied how telomere length links spatiotemporal variation in early-life conditions to survival in the Seychelles warbler (*Acrocephalus sechellensis*). The Seychelles warbler system is excellently suited to addressing this question, due to the excellent, long-term lonigtudinal data on survival and senescence. We combined this with cross-seciotnal and longitudinal telomere measurements taken from over 300 birds, and analyse how the variation in the early life environment within and across cohorts affects telomere length and survival.

4. We show that early-life telomere length varies markedly across cohorts, and that this variation is positively related to annual variation in food availability. We then show that having short telomeres as a juvenile is associated with reduced late life survival, but that this relationship was subject to strong cohort effects. Having short telomeres was associated with reduced late-life survival among individuals born in years where food was limited, but not among individuals born in years with high food availability.

5. Our results show that telomeres can reveal hidden costs at the individual and cohort levels, and suggest that they may provide a link between early-life conditions and late-life survival in wild populations.

@@

###Introduction
Being subject to favourable environmental conditions during development and growth can confer fitness advantages later in life (so called 'silver spoon effects') [@Grafen1988; @Monaghan2008]. There is now evidence that adult fitness can be affected by a range of early-life experiences, including variation in habitat quality and resource availability [@Madsen2000; @VandePol2006; @Hayward2013], population density [@Nussey2007;@Douhard2013] and natural or anthropogenic environmental disturbance [@Reid2003; @Cartwright2014]. Understanding the causes and consequences of these early-life experiences is key to understanding many ecological and evolutionary  processes, including  patterns of natural and sexual selection,  population growth rates and even local extinction [@Coulson2001; @Roach2014]. Understanding silver spoon effects is therefore of central interest to ecologists, evolutionary biologists and conservationists.

@@

The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent model system for studying senescnece in the wild [@Hammers2015]. Specifically, it is ideally suited for assessing the relationship between early-life environmental variation, telomere length and late-life fitness. Firstly, due to the isolated nature of the study population and intensive field monitoring, we have excellent survival data and sampling material spanning many years (see methods, below). Secondly, telomere length predicts survival independently of age in adult Seychelles warblers, suggesting that telomeres act as a biomarker of cost in this species [@Barrett2013a]. Finally, ecological conditions on Cousin are variable over space and time due to spatiotemporal variation in foliage cover and food availability [@VandeCrommenacker2011a].

@@

@@

###Acknowledgements
Nature Seychelles facilitate the long-term Seychelles warbler project, allowing us to stay on Cousin Island, and providing accommodation and facilities during field work. The Seychelles Bureau of Standards and Department of Environment gave permission for sampling and fieldwork. Emma Barrett laid the foundations for this study, generating much of the telomere data. We thank everyone who has helped in the field, and the current Seychelles warbler research group for useful discussions. This work was funded by two Natural Environment Research Council (NERC) grants to DSR (NE/F02083X/1 and NE/K005502/1) on which JK, TB and HLD were project partners. HLD was funded by a NERC fellowship.
